---
title: "Interpretation, analysis and Customer segmentation"
output: github_document
---
```{r}
load(".RData")
packages <- readLines("requirements.txt")
lapply(packages, library, character.only = TRUE)
```

# Clustering algorithm
The model that I chose to describe the customer segmentation is CLARA with 6 clusters. This decision was taken after careful analysis and comparison between other clustering algorithms.

### Visualization of the clusters
```{r}
fviz_cluster(clara_res1,
             data=scaled_behav,
             ellipse.type = "euclid",
             repel = FALSE,
             ggtheme = theme_classic(),
             palette = 'jco',
             main = "CLARA 6 CLUSTERS")
```
**The first step is to add the 'cluster' column', which describes the cluster affiliations of each data object.**
```{r}
clust_data <- cbind(data, clara_res1$clustering)
colnames(clust_data)[ncol(clust_data)]<-"Clusters" #Last column
head(clust_data)
```

### Number of objects in each cluster
```{r}
clust_data %>%
  group_by(Clusters) %>%
  summarise(count = n())
```

## Interpretation of the clusters
Descriptive statistics for each feature relative to cluster membership. Examination of arithmetic mean, maximum, and minimum.
```{r}
clust_summary <- clust_data %>%
  group_by(Clusters) %>%
  summarise(across(where(is.numeric),
                   list(mean = mean, min = min, max = max),
                   .names = "{.col}_{.fn}")) %>% ungroup()

clust_summary
```

First, I will look at several indicators related to customer behavior at the bank. I want to look at some basic ones such as mortgages, monthly expenses, loan amounts, and stock investments.

I will also look at indicators such as the number of logins to the app, time spent online in the system, Bitcoins, NFTs, and the number of social media channels related to the fintech industry and new virtual currencies. I believe that the indicators for these signs would be higher for customers who would be interested in a service such as WalletUp. WalletUp's main focus is on online banking services, which would primarily interest customers who spend time online and are interested in modern solutions in the financial industry.

## Analysis of mean values for the main financial behavioral factors
```{r}
ggplot(data = clust_summary, aes(x = factor(Clusters), y = clust_summary$MortageVolume_mean)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = scales::comma)+
  labs(x = "Cluster", y = "Mean Mortgage Volume") +
  ggtitle("Mean Mortgage Volume by Cluster")+
  theme_minimal()

ggplot(data = clust_summary, aes(x = factor(Clusters), y = clust_summary$AccountSpending_mean)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = scales::comma)+
  labs(x = "Cluster", y = "Mean Account Expenditures") +
  ggtitle("Mean Monthly Expenditures by Cluster")+
  theme_minimal()

ggplot(data = clust_summary, aes(x = factor(Clusters), y = clust_summary$CreditVolume_mean)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = scales::comma)+
  labs(x = "Cluster", y = "Mean Credit Volume") +
  ggtitle("Mean Credit Volume Cluster")+
  theme_minimal()

ggplot(data = clust_summary, aes(x = factor(Clusters), y = clust_summary$StockVolume_mean)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = scales::comma)+
  labs(x = "Cluster", y = "Mean Investments in Stocks") +
  ggtitle("Mean Stocks Investments by Cluster")+
  theme_minimal()
```

At this point, we can draw the following conclusions: Cluster 1 has an average mortgage of $300k, but the other three indicators are low. Cluster 2 has a great average monthly credit expense of $2,500, while all other characteristics are in last place. Cluster 4 has the largest credit expense, but also the highest investment in stocks. Customers in cluster 5 have relatively steep monthly expenses than the other five clusters ($3,500). Cluster number 6 pays the highest mortgage of around $400k, but their other expenses are medium compared to the other segments. Cluster 3 is the most balanced, with rising investments. ($3,500).

```{r}
ggplot(data = clust_data, aes(x = factor(Clusters), y = AppLogins)) +
  geom_boxplot() +
  labs(x = "Clusters", y = "App Logins", title = "Boxplot of App Logins by Cluster") +
  theme_minimal()

ggplot(data = clust_data, aes(x = factor(Clusters), y = TimeOnlineBanking)) +
  geom_boxplot() +
  labs(x = "Clusters", y = "Time Spent in Online Banking (Min)", title = "Boxplot of Time Spent Online by Cluster") +
  theme_minimal()

ggplot(data = clust_data, aes(x = factor(Clusters), y = Bitcoins)) +
  geom_boxplot() +
  labs(x = "Clusters", y = "Bitcoins", title = "Boxplot of Bitcoins Ownership by Cluster") +
  theme_minimal()

ggplot(data = clust_data, aes(x = factor(Clusters), y = NFTs)) +
  geom_boxplot() +
  labs(x = "Clusters", y = "NFTs", title = "Boxplot of NFTs Ownership by Cluster") +
  theme_minimal()

ggplot(data = clust_data, aes(x = factor(Clusters), y = SocialMediaInter)) +
  geom_boxplot() +
  labs(x = "Clusters", y = "Social Media Channels followed", title = "Boxplot of Followed Financial Social Media Channels by Cluster") +
  theme_minimal()

ggplot(data = clust_data, aes(x = factor(Clusters), y = BranchVisits)) +
  geom_boxplot() +
  labs(x = "Clusters", y = "Branch Visits", title = "Boxplot of Branch visits by Cluster") +
  theme_minimal()
```
At this stage, we can draw the following conclusions: Cluster 5, which racks up the highest monthly expenses, actually has the fewest virtual assets and doesn’t really spend time on online banking. Cluster 4, on the other hand, is definitely more risky in terms of loans and investments, but this group is also diving into digital assets, holding around 0.5 Bitcoins and in some cases more than 10 NFTs. They also spend the most time in online banking and following finance-related social media channels.

Cluster 3 looks quite similar to Cluster 4, just one step lower on almost every factor—except Bitcoin ownership, where they actually rank at the top. Clusters 1 and 2 show a solid interest in the fintech space, as we see more clients in these groups following social media finance channels than the cluster average. They spend a decent amount of time on online banking, but they’re way more cautious when it comes to investing in things like Bitcoin and NFTs. Finally, Cluster 6 is heavy on online banking, invests in Bitcoin, but tends to follow smaller, more niche finance channels.

When it comes to visiting the bank in person, Clusters 1, 2, 3, and 4 rarely go to branches, while Cluster 5 shows up pretty often (10 times over the past 8 weeks). Cluster 6 also pops in about 2–3 times a month.

Based on this breakdown, I’d aim to build targeted marketing campaigns for Clusters 3, 4, 1, and 6. That’s because 3, 4, and 6 are clearly into online services, while Cluster 1 is massive—covering 30% of the whole dataset. To get a better understanding of all these groups, we’ll also take a closer look at their socio-demographic traits.

## Analysis of mean values of socio-demographic factors

### Age
```{r}
ggplot(data = clust_data, aes(x = factor(Clusters), y = Age)) +
  geom_boxplot() +
  labs(x = "Clusters", y = "Age", title = "Boxplot of Age by Cluster") +
  theme_minimal()
```
Starting with the age of our customers, it becomes clear that customers in cluster number 5 are older, with an average age of almost 60. Clusters 3 and 4 have an average age of around 40, but in cluster 3 we have customers who are close to the maximum and minimum ages, i.e. high dispersion. The interesting thing about the largest cluster, cluster 1, is that we have an average age of 25, which is low compared to the others. 50% of cluster 1 are under 45 years of age. Cluster 2 consists mainly of young consumers with some exceptions, and cluster 6 has an average age of 30.

### Income
```{r}
ggplot(data = clust_summary, aes(x = factor(Clusters), y = clust_summary$Income_mean)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = scales::comma)+
  labs(x = "Cluster", y = "Mean Income") +
  ggtitle("Mean Income by Cluster")+
  theme_minimal()
```
If we examine the monthly income of the clusters obtained, we can see that clusters 1, 3, and 4 are very close ($75k). Cluster 5 has the highest average income of $175k, which leads to a positive linear correlation with the age of customers in this cluster and justifies their high monthly expenses. In second place is cluster 6, which is one of the segments with younger customers who spend time online.

### Household size
```{r}
ggplot(data = clust_data, aes(x = factor(Clusters), y = HouseholdSize)) +
  geom_boxplot() +
  labs(x = "Clusters", y = "HouseholdSize", title = "Boxplot of Household Size by Cluster") +
  theme_minimal()
```
Since segments 1, 3, 4, and 6 consist mainly of young customers, we can see that they mostly live with their partner without children or, most often, with one child.

### Internet traffic
```{r}
ggplot(data = clust_summary, aes(x = factor(Clusters), y = clust_summary$InternetTrafficVolume_mean)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = scales::comma)+
  labs(x = "Cluster", y = "Mean Internet traffic") +
  ggtitle("Mean Internet traffic by Cluster")+
  theme_minimal()
```
Clusters 3 and 4, which we knew spent a relatively large amount of time banking online and followed quite a few fintech channels on social media, can be seen to have high overall internet traffic. Our older customers do not spend time online and are not suitable for our campaign, while clusters 1 and 6 have a stable average Internet traffic of 50 GB.

## Customer Personas
#### As table:
```{r}
personas <- data.frame(
  Persona = c("Risky Investor", 
              "Innovative Investor", 
              "Established & Open", 
              "Young Professional", 
              "The Indebted", 
              "Mature Traditionalist"),
  Cluster = c(4, 3, 6, 1, 2, 5),
  Age_Median = c(38, 42, 29, 25, 23, 58),
  Income_Median = c(75000, 70000, 130000, 60000, 40000, 175000),
  Household_Size = c("With partner (2)", 
                     "Partner + 1 child (3)", 
                     "Partner + child (3)", 
                     "With partner (2)", 
                     "Partner + child (3)", 
                     "Partner + 2 children (4)"),
  Behaviour = c("High-risk profile due to loans and investments. Prefers Bitcoins/NFTs. Very high online activity.",
                "Active online, follows social channels. Owns Bitcoins/NFTs. Balanced loans/expenses vs. investments.",
                "High mortgage, medium investment value. Good online banking activity. Owns Bitcoins.",
                "Moderate online activity and interest in fintech. Low spending/loans, but relatively high mortgage. Low interest in digital assets.",
                "High loans, low investments. Moderate-to-low online activity. Financial instability.",
                "Very low online activity. High income but also high monthly expenses. Weak asset investing."),
  Potential = c("High", "High", "Medium–High", "Medium", "Low", "Very Low"),
  Comments = c("Early adopters. 1,250 clients.",
               "Tech-advanced group. Lower risk than Cluster 4. 1,000 clients.",
               "Needs education on WalletUp benefits. High potential for investing in WalletUp. 1,500 clients.",
               "Large 'middle-class' base. Could be marketed app benefits beyond digital assets. 2,959 clients (largest group).",
               "No interest in investing. 2,541 clients. Not a priority.",
               "Conservative, older clients. Prefers traditional banking. 1,500 clients.")
)

print(personas)
```

# Conclusion

## Marketing strategy
After running a detailed cluster analysis, my first recommendation is to focus on the “Risky Investor” and the “Innovative Investor.” These two groups spend the most time online and would adapt quickly to using the app. The Risky Investor, in particular, makes bold financial moves and is more likely to experiment with new investment methods, making them an ideal early adopter.

The Established & Open group (Cluster 6) also shows strong potential. They’re active online but currently focus more on real estate through mortgages. With the right guidance and education, this segment could be encouraged to diversify. Their higher average income also makes them a valuable and profitable target. At the same time, seeing peers from Clusters 3 and 4 adopt the app could motivate them to take action.

The Young Professionals represent the largest customer base. They have a decent online presence, and with the right campaign, WalletUp could draw them in by showing different investment opportunities beyond just traditional assets. Their large numbers open the door for testing and scaling multiple marketing strategies.

On the other hand, the Indebted group isn’t a good fit right now—they’re financially unstable and less engaged online compared to the other clusters.

Lastly, the Mature Traditionalists are high-income earners but older and less digitally active. They prefer visiting branches in person. While most campaigns should be digital (especially via social media), this group is best approached offline through branch staff interactions or printed materials like flyers.